Bow shocks, bow waves, and dust waves. III. Diagnostics

### 2019-04-01 papers/Papers1234/bs-bw-dw-03.tex
Submit Paper III to MNRAS and astro-ph

Bow-shaped circumstellar nebulae are observed around a wide variety of
stars \citep{Gull:1979a, Cox:2012a, Cordes:1993a} but are most
numerous around luminous OB stars \citep{Kobulnicky:2016a}.  They are
most commonly interpreted as bow shocks, due to a supersonic relative
motion of the surrounding medium, which interacts with the stellar
wind \citep{Wilkin:1996a}.  Early surveys for bow shocks around OB
stars \citep{van-Buren:1995a} concentrated on runaway stars
\citep{Blaauw:1961a, Hoogerwerf:2001a} that have been ejected at high
velocities (\(> \SI{30}{km.s^{-1}}\)) due to multi-body encounters in
star clusters or due to the core-collapse supernova explosion of a
binary companion.  However, only a small fraction of runaways show
detectable bow shocks \citep{Huthoff:2002a, Peri:2012a, Peri:2015a,
  Prisegen:2019a}. Targeted searches at mid-infrared wavelengths of
particular high-mass star forming regions such as M17 and RCW~49
\citep{Povich:2008a}, Cygnus~X \citep{Kobulnicky:2010a}, and Carina
\citep{Sexton:2015b} have revealed many bows around slower moving
stars.  In many cases, it may be streaming motions in the interstellar
medium that provide most of the relative velocity: weather vanes
rather than runaways \citep{Povich:2008a}.  More recently, large-scale
surveys of the Galactic Plane \citep{Kobulnicky:2016a,
  Kobulnicky:2017a} and of nearby stars in the Bright Star Catalog
\citep{Bodensteiner:2018a} have revealed hundreds more such bow shocks.

Analytic and semi-analytic thin-shell models of bow shocks have been
developed \citep{van-Buren:1992a, Wilkin:1996a, Canto:1996},
including the effects of non-spherical winds and nonaxisymmetric bows
\citep{Wilkin:2000a, Henney:2002a, Canto:2005a, Tarango-Yong:2018a}.
Increasingly realistic numerical hydrodynamic simulations have been
performed \citep{Matsuda:1989a, Raga:1997a, Comeron:1998a,
  Meyer:2014b, Mackey:2015a}, including magnetic fields
\citep{Meyer:2017a, Katushkina:2017a, Katushkina:2018a} and detailed
predictions of the dust emission \citep{Meyer:2016a, Acreman:2016a,
  Mackey:2016a}.

@@

We have proposed a novel diagnostic method, the \(\tau\)-\(\eta\shell\)
diagram, for analyzing observations of stellar bow shocks around OB
stars, which allows discrimination between radiation-supported and
wind-supported bows.  Our principal results are as follows:
\begin{enumerate}[1.]
\item The UV optical depth \(\tau\) of the bow shell can be estimated
  from the observed infrared shell luminosity \(L\IR\) and stellar
  luminosity \(L_*\) as \(\tau = 2 L\IR/L_*\)
  (\S~\ref{sec:energy-trapp-vers}).
\item The shell momentum efficiency \(\eta\shell\), which is the
  fraction of stellar radiative momentum that is transferred to the
  shell either directly or indirectly (via stellar wind), can be
  estimated from \(\tau\) and the observed shell radius \(R_0\), after
  making some auxiliary assumptions about the physical conditions in
  the shell (\S~\ref{sec:energy-trapp-vers}).
\item By comparing \(\tau\) and \(\eta\shell\), it is possible to
  discriminate between three regimes (\S~\ref{sec:eta-tau-diagnostic}
  and see Paper~I): wind-supported bow shocks
  (\(\eta\shell \gg \tau\)), radiation-supported bow waves
  (\(\eta\shell \sim \tau < 1\)), and radiation-supported bow shocks
  (\(\eta\shell \sim \tau > 1\)).
\item Shells with \(\eta\shell < \num{e-3}\) are potentially in a
  fourth regime of dust waves, where gas--grain collisional coupling
  breaks down (see Paper~II).
\item By analyzing a published sample of 20 mid-infrared bow shock
  candidates \citep[K18]{Kobulnicky:2018a}, plus additional compact
  bows in Orion, we find 4 strong candidates for radiation-supported
  bow waves plus 3 marginal cases (\S\S~\ref{sec:cand-radi-supp} and
  \ref{sec:comm-prop-cand}).  No strong candidates for dust waves are
  found.
\item For wind-supported bow shocks, the stellar wind mass-loss rate
  can be found from \(\tau\) and \(\eta\shell\)
  (\S~\ref{sec:mass-loss-determ}). We compare this method with the
  previously proposed method of K18 (\S~\ref{app:bow-shock-data}) and
  suggest a correction to the dust emissivities used in the latter,
  which reduces the mass-loss rates by a factor of about 2.
\item After this correction, the two mass-loss methods agree well for
  small bows (\(R_0 \le \SI{0.3}{pc}\)), but the
  \(\tau\)--\(\eta\shell\) method under-predicts the mass-loss rate by
  up to a factor of 10 for larger bows, which also tend to have
  relatively thin shells (\S~\ref{sec:underst-diff-betw}).  This
  discrepancy could be eliminated by extending the
  \(\tau\)--\(\eta\shell\) method to include the observed shell
  thickness.
\item Disruption of the largest dust grains by rotational torques in
  the strong circumstellar radiation field is predicted to have a
  relatively modest effect on the optical properties of the dust
  mixture (\S~\ref{sec:evolution-grain-size}), increasing the UV
  opacity by about 50\%.
\end{enumerate}

>>>>

Bow-shaped circumstellar nebulae are observed around a wide variety of
stars \citep{Gull:1979a, Cox:2012a, Cordes:1993a} but are most
numerous around luminous OB stars \citep{Kobulnicky:2016a}.  They are
most commonly interpreted as bow shocks, due to a supersonic relative
motion of the surrounding medium, which interacts with the stellar
wind \citep{Wilkin:1996a}.  Early surveys for bow shocks around OB
stars \citep{van-Buren:1995a} concentrated on runaway stars
\citep{Blaauw:1961a, Hoogerwerf:2001a} that have been ejected at high
velocities (\(> \SI{30}{km.s^{-1}}\)) due to multi-body encounters in
star clusters or due to the core-collapse supernova explosion of a
binary companion.  However, only a small fraction of runaways show
detectable bow shocks \citep{Huthoff:2002a, Peri:2012a, Peri:2015a,
  Prisegen:2019a}. Targeted searches at mid-infrared wavelengths of
particular high-mass star forming regions such as M17 and RCW~49
\citep{Povich:2008a}, Cygnus~X \citep{Kobulnicky:2010a}, and Carina
\citep{Sexton:2015b} have revealed many bows around slower moving
stars.  In many cases, it may be streaming motions in the interstellar
medium that provide most of the relative velocity: weather vanes
rather than runaways \citep{Povich:2008a}.  More recently, large-scale
surveys of the Galactic plane \citep{Kobulnicky:2016a,
  Kobulnicky:2017a} and of nearby stars in the Bright Star Catalog
\citep{Bodensteiner:2018a} have revealed hundreds more such bow shocks.

Analytic and semi-analytic thin-shell models of bow shocks have been
developed \citep{van-Buren:1992a, Wilkin:1996a, Canto:1996},
including the effects of non-spherical winds and nonaxisymmetric bows
\citep{Wilkin:2000a, Henney:2002a, Canto:2005a, Tarango-Yong:2018a}.
Increasingly realistic numerical hydrodynamic simulations have been
performed \citep{Matsuda:1989a, Raga:1997a, Comeron:1998a, Arthur:2006a,
  Meyer:2014b, Mackey:2015a}, including magnetic fields
\citep{Meyer:2017a, Katushkina:2017a, Katushkina:2018a} and detailed
predictions of the dust emission \citep{Meyer:2016a, Acreman:2016a,
  Mackey:2016a}.

@@

We have proposed a novel diagnostic method, the \(\tau\)-\(\eta\shell\)
diagram, for analyzing observations of stellar bow shocks around OB
stars, which allows discrimination between radiation-supported and
wind-supported bows.  Our principal results are as follows:
\begin{enumerate}[1.]
\item The UV optical depth \(\tau\) of the bow shell can be estimated
  from the observed infrared shell luminosity \(L\IR\) and stellar
  luminosity \(L_*\) as \(\tau = 2 L\IR/L_*\)
  (\S~\ref{sec:energy-trapp-vers}).
\item The shell momentum efficiency \(\eta\shell\), which is the
  fraction of stellar radiative momentum that is transferred to the
  shell either directly or indirectly (via stellar wind), can be
  estimated from \(\tau\) and the observed shell radius \(R_0\), after
  making some auxiliary assumptions about the physical conditions in
  the shell (\S~\ref{sec:energy-trapp-vers}).
\item By comparing \(\tau\) and \(\eta\shell\), it is possible to
  discriminate between three regimes (\S~\ref{sec:eta-tau-diagnostic}
  and see Paper~I): wind-supported bow shocks
  (\(\eta\shell \gg \tau\)), radiation-supported bow waves
  (\(\eta\shell \sim \tau < 1\)), and radiation-supported bow shocks
  (\(\eta\shell \sim \tau > 1\)).
\item Shells with \(\eta\shell < \num{e-3}\) are potentially in a
  fourth regime of dust waves, where gas--grain collisional coupling
  breaks down (see Paper~II).
\item By analyzing a published sample of 20 mid-infrared bow shock
  candidates \citep[K18]{Kobulnicky:2018a}, plus additional compact
  bows in Orion, we find 4 strong candidates for radiation-supported
  bow waves plus 3 marginal cases (\S\S~\ref{sec:cand-radi-supp} and
  \ref{sec:comm-prop-cand}).  No strong candidates for dust waves are
  found.
\item For wind-supported bow shocks, the stellar wind mass-loss rate
  can be found from \(\tau\) and \(\eta\shell\)
  (\S~\ref{sec:mass-loss-determ}). We compare this method with the
  previously proposed method of K18 (\S~\ref{app:bow-shock-data}) and
  suggest a correction to the dust emissivities used in the latter,
  which reduces the mass-loss rates by a factor of about 2.
\item After this correction, the two mass-loss methods agree well for
  small bows (\(R_0 \le \SI{0.3}{pc}\)), but the
  \(\tau\)--\(\eta\shell\) method under-predicts the mass-loss rate by
  up to a factor of 10 for larger bows, which also tend to have
  relatively thin shells (\S~\ref{sec:underst-diff-betw}).  This
  discrepancy could be eliminated by extending the
  \(\tau\)--\(\eta\shell\) method to include the observed shell
  thickness.
\item Disruption of the largest dust grains by rotational torques in
  the strong circumstellar radiation field is predicted to have a
  relatively modest effect on the optical properties of the dust
  mixture (\S~\ref{sec:evolution-grain-size}), increasing the UV
  opacity by about 50\%.
\item The bow source KGK2010~2 is a low-luminosity B~star with an
  anomalously high apparent mass-loss rate.  This discrepancy could be
  eliminated if the bow shell has trapped the ionization front and is
  predominantly neutral, in which case it would become yet another
  candidate for radiation support.
\end{enumerate}

### 2019-04-01 papers/Papers1234/sec-observational-diagnostics.tex
Submit Paper III to MNRAS and astro-ph

Therefore, we have shown that source~342 cannot have a neutral shell
if we take the published observational data at face value.  However,
more recent spectroscopic observations revise the spectral
classification to B4~V (H.~Kobulnicky, priv.~comm.) and this would
change the picture completely.  A B4 dwarf, with effective temperature
of \SI{16700}{K} \citep[Tab.~4 of][]{Pecaut:2013a} would have a very
small ionizing luminosity. Interpolating on Table~4 of
\citet{Lanz:2007a}, we find \(S_{49} / L_4 \approx \num{4e-6}\), which
yields \(\tau\iftrap = 0.017\) from equation~\eqref{eq:tau-trap-rbw}.
At the same time, the revised spectroscopy implies a smaller \(L_*\)
by roughly a factor of 10, which increases the shell's derived
\(\tau\) by the same factor (eq.~[\ref{eq:tau-empirical}]), yielding
\(\tau \approx 0.05\).  Since we now have \(\tau > \tau\iftrap\) it is
possible that ionization front may be trapped, giving a predominately
neutral shell.  This also relies, however, on the radiation field from
the nearby Cygnus OB2 cluster being insufficient to ionize the shell.
Given that the cluster's stellar population \citep{Wright:2015a}
includes 3 WR stars and a handful of early-O supergiants, the ionizing
luminosity must be of order \SI{e51}{s^{-1}}.  In the absence of
absorption, and assuming that the true separation to be no more than a
few times larger than the projected separation, the cluster stars
would contribute over 100 times greater ionizing flux at the position
of source 34 than that due to the B4 star itself.  It is therefore
necessary for the source to lie outside of the cluster's \hii{} region
in order for the shell to be neutral.  It is quite plausible that this
might be the case, since radio continuum observations \citetext{Fig.~4
  of \citealp{Wendker:1991a} and Fig.~7 of \citealp{Tung:2017a}}
indicate that the neighborhood of source~342 is a local minimum in
free-free emission between discrete \hii{} regions of the Cygnus~X
complex, while Herschel maps \citep{Schneider:2016a} show the
presence of both warm and cold dust along the line of sight and clumps
of Class~0 and~I YSOs are seen nearby \citep{Beerer:2010a}, indicative
of shielded neutral/molecular gas.

>>>>

Therefore, we have shown that source~342 cannot have a neutral shell
if we take the published observational data at face value.  However,
more recent spectroscopic observations revise the spectral
classification to B4~V (H.~Kobulnicky, priv.~comm.) and this would
change the picture completely.  A B4 dwarf, with effective temperature
of \SI{16700}{K} \citep[Tab.~4 of][]{Pecaut:2013a} would have a very
small ionizing luminosity. Interpolating on Table~4 of
\citet{Lanz:2007a}, we find \(S_{49} / L_4 \approx \num{4e-6}\), which
yields \(\tau\iftrap = 0.017\) from equation~\eqref{eq:tau-trap-rbw}.
At the same time, the revised spectroscopy implies a smaller \(L_*\)
by roughly a factor of 10, which increases the shell's derived
\(\tau\) by the same factor (eq.~[\ref{eq:tau-empirical}]), yielding
\(\tau \approx 0.05\).  Since we now have \(\tau > \tau\iftrap\) it is
possible that ionization front may be trapped, giving a predominately
neutral shell.  This also relies, however, on the radiation field from
the nearby Cygnus OB2 cluster being insufficient to ionize the shell.
Given that the cluster's stellar population \citep{Wright:2015a}
includes 3 WR stars and a handful of early-O supergiants, the ionizing
luminosity must be of order \SI{e51}{s^{-1}}.  In the absence of
absorption, and assuming that the true separation be no more than a
few times larger than the projected separation, the cluster stars
would contribute over 100 times greater ionizing flux at the position
of source~342 than that due to the B4 star itself.  It is therefore
necessary for the source to lie outside of the cluster's \hii{} region
in order for the shell to be neutral.  It is quite plausible that this
might be the case, since radio continuum observations \citetext{Fig.~4
  of \citealp{Wendker:1991a} and Fig.~7 of \citealp{Tung:2017a}}
indicate that the neighborhood of source~342 is a local minimum in
free-free emission between discrete \hii{} regions of the Cygnus~X
complex, while Herschel maps \citep{Schneider:2016a} show the
presence of both warm and cold dust along the line of sight and clumps
of Class~0 and~I YSOs are seen nearby \citep{Beerer:2010a}, indicative
of shielded neutral/molecular gas.
